Enforced Invariant Confluence

### 2016-11-26 doc/counters.tex
Crunched figure.
Added two more proofs.

\begin{theorem}\label{thm:one-is-enough}
  For all sets of transactions $T$ and invariants $I$, T is 1-\iconfluent{} if
  and only if it is \iconfluent{}.
\end{theorem}
\begin{proof}
  The reverse direction is immediate from the definition of \iconfluence{}. For
  the forward direction, consider a set of transactions $T$ that is
  1-\iconfluent{} with respect to an invariant $I$, but assume for
  contradiction that $T$ is not \iconfluent{} with respect to $I$. That is,
  there exists a database state $D$ and two chains $C^t = t_1, t_2, \ldots,
  t_m$ and $C^u = u_1, u_2, \ldots, u_n$ such that $D$ satisfies $I$,
  $C^t$ satisfies $I$ starting at $D$, and $C^u$ satisfies $I$ starting at $D$,
  but $D \circ C^t \circ C^u$ does not satisfy $I$.

@@

  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.8]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
    \end{tikzpicture}
    \caption{Visualization of $I$.}
    \label{fig:wimp-iviz}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.8]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxy{\draw[vec] (\x, \y) -- (\x, \y + 1);}
    \end{tikzpicture}
    \caption{Visualization of $c_1$.}
    \label{fig:wimp-c1viz}
  \end{subfigure}

  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.8]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxysmall{\draw[vec] (\x, \y) -- (\x + \x + 1, \y);}
    \end{tikzpicture}
    \caption{Visualization of $c_2$.}
    \label{fig:wimp-c2viz}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.8]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxysmall{\draw[vec] (\x, \y) -- (\x + \x - 1, \y);}
    \end{tikzpicture}
    \caption{Visualization of $c_3$.}
    \label{fig:wimp-c3viz}
  \end{subfigure}

@@

\subsection{\iconfluence{} Alternatives}
\begin{itemize}
  \item \todo{\istrengthstar $\centernot \implies$ \iconfluence?}
  \item \todo{\iconfluence $\implies$ \istrengthstar?}
  \item \todo{Find something between \isafety{} and \iconfluence.}
  \item \todo{1-iconfluence with derived txns to start}
\end{itemize}

@@

  \item \textbf{\ipreservation.} $\forall D, c.\>
       I(D) \implies
       I(D \circ \denote{c}(D)$

@@

  \item \textbf{\istrengthstar{} \cite{gotsman2016cause}.}
    \leqnomode
    \begin{flalign*}
      & \exists G_0 \subseteq (\var \to \ints) \times (\var \to \ints).& \\
      & G_0(I) \subseteq I \land{} & \tag*{S2.} \\
      & \forall c, D, D'.\>
          D \in I \land
          (D, D') \in G_0^* \implies
          (D', \denote{c}(D)(D')) \in G_0 & \tag*{S3.}
    \end{flalign*}
    \reqnomode
    where $I$ is viewed as a set of database states.

  \item \textbf{\istrength{} \cite{gotsman2016cause}.}
    \leqnomode
    \begin{flalign*}
      & \exists G_0 \subseteq (\var \to \ints) \times (\var \to \ints). & \\
      & G_0(I) \subseteq I \land{} & \tag*{T2.} \\
      & \forall c.\> & \tag*{T3.} \\
      & \exists P_1, \ldots, P_n \subseteq (\var \to \ints).\> & \\
      & \exists Q_1, \ldots, Q_n \subseteq (\var \to \ints).\> & \\
      & \quad I \subseteq \cup_{i=1}^n P_i \land {} & \tag*{T3b.} \\
      & \quad \forall i \in \set{1, \ldots, n}.\> & \\
      & \quad \quad P_i \subseteq Q_i \land {} & \tag*{T3c.} \\
      & \quad \quad G_0(Q_i) \subseteq Q_i \land {} & \tag*{T3d.} \\
      & \quad \quad Q_i \times \denote{c}(P_i)(Q_i) \subseteq G_0 & \tag*{T3e.}
    \end{flalign*}
    \reqnomode
    where $I$ is viewed as a set of database states.
\end{itemize}

\begin{claim}
  0-\isafety $\implies$ \istrengthstar.
\end{claim}
\begin{proof}
  Consider an abitrary set of \imp{} transactions $T$ that is 0-\isafe{} with
  respect to invariant $I$. We show $T$ is \istrongstar{} with respect to $I$.
  Let $G_0 = I \times I$. S2 holds trivially: $G_0(I) = I
  \subseteq I$. Consider an arbitrary $c$, $D$, and $D'$ where $D \in I$ and
  $(D, D') \in G_0^*$.
  \begin{align*}
    & D \in I \land (D, D') \in G_0^*  \\
    &\implies D \in I \land D' \in I \tag{Definition of $G_0$} \\
    &\implies D' \in I \land D' \circ \denote{c}(D) \in I \tag{0-\isafety} \\
    &\implies (D', D' \circ \denote{c}(D)) \in G_0 \tag{Definition of $G_0$}
  \end{align*}
\end{proof}

\begin{claim}
  \istrengthstar $\centernot\implies$ 0-\isafety
\end{claim}
\begin{proof}
  Consider a database over a single variable $x$. For convenience, we'll denote
  database states as integers. Let
  \begin{itemize}
    \item
      $G \defeq \set{(0, 0), (0, 2)} \cup \setst{(a, b)}{a \geq 2, b \geq a}$
    \item
      $I \defeq x = 0 \lor x \geq 2$, and
    \item
      the single \imp{} transaction $c \defeq
      \impif{read(x)=0}{add(x, 0)}{add(x, 1)}$.
  \end{itemize}

@@

>>>>

\begin{theorem}\label{thm:one-is-enough}
  For all sets of transactions $T$ and invariants $I$, T is 1-\iconfluent{} if
  and only if it is \iconfluent{}.
\end{theorem}
\begin{proof}
  The reverse direction is immediate from the definition of \iconfluence{}. For
  the forward direction, consider a set of transactions $T$ that is
  1-\iconfluent{} with respect to an invariant $I$. Consider an arbitrary
  database state $D$ and two chains $C^t = t_1, t_2, \ldots, t_m$ and $C^u =
  u_1, u_2, \ldots, u_n$ such that $D$ satisfies $I$, $C^t$ satisfies $I$
  starting at $D$, and $C^u$ satisfies $I$ starting at $D$. We show $D \circ
  C^t \circ C^u$ satisfies $I$.

@@

\begin{figure}[h]
  \centering
  \newcommand{\thescale}{0.7}

  \begin{subfigure}[b]{0.24\textwidth}
    \centering
    \begin{tikzpicture}[scale=\thescale]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
    \end{tikzpicture}
    \caption{Visualization of $I$.}
    \label{fig:wimp-iviz}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \centering
    \begin{tikzpicture}[scale=\thescale]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxy{\draw[vec] (\x, \y) -- (\x, \y + 1);}
    \end{tikzpicture}
    \caption{Visualization of $c_1$.}
    \label{fig:wimp-c1viz}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \centering
    \begin{tikzpicture}[scale=\thescale]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxysmall{\draw[vec] (\x, \y) -- (\x + \x + 1, \y);}
    \end{tikzpicture}
    \caption{Visualization of $c_2$.}
    \label{fig:wimp-c2viz}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \centering
    \begin{tikzpicture}[scale=\thescale]
      \gridxy{}
      \forxy{\node[point] (\x-\y) at (\x, \y) {};}
      \forxysmall{\draw[vec] (\x, \y) -- (\x + \x - 1, \y);}
    \end{tikzpicture}
    \caption{Visualization of $c_3$.}
    \label{fig:wimp-c3viz}
  \end{subfigure}

@@

\subsection{\iconfluence{} Alternatives}
\begin{itemize}
  \item \todo{\istrengthstar $\centernot \implies$ \iconfluence?}
  \item \todo{\iconfluence $\implies$ \istrengthstar?}
  \item \todo{\istrength $\centernot \implies$ \isafety?}
  \item \todo{\isafety $\implies$ \istrength?}
  \item \todo{\istrength $\centernot \implies$ \iconfluence?}
  \item \todo{\iconfluence $\implies$ \istrength?}
  \item \todo{Find something between \isafety{} and \iconfluence.}
  \item \todo{1-iconfluence with derived txns to start}
\end{itemize}

@@

  \item \textbf{\ipreservation.} $\forall D, c.\>
       I(D) \implies
       I(D \circ \denote{c}(D))$

@@

  \item \textbf{\istrengthstar{} \cite{gotsman2016cause}.}
    \leqnomode
    \begin{flalign*}
      & \exists G_0 \subseteq (\var \to \ints) \times (\var \to \ints).& \\
      & G_0(I) \subseteq I \land{} & \tag*{S2.} \\
      & \forall c, D, D'.\>
          D \in I \land
          (D, D') \in G_0^* \implies
          (D', D' \circ \denote{c}(D)) \in G_0 & \tag*{S3.}
    \end{flalign*}
    \reqnomode
    where $I$ is viewed as a set of database states.

  \item \textbf{\istrength{} \cite{gotsman2016cause}.}
    \leqnomode
    \begin{flalign*}
      & \exists G_0 \subseteq (\var \to \ints) \times (\var \to \ints). & \\
      & G_0(I) \subseteq I \land{} & \tag*{T2.} \\
      & \forall c.\>
        \exists P_1, \ldots, P_n \subseteq (\var \to \ints).\>
        \exists Q_1, \ldots, Q_n \subseteq (\var \to \ints).\> & \tag*{T3.} \\
      & \quad I \subseteq \cup_{i=1}^n P_i \land {} & \tag*{T3b.} \\
      & \quad \forall i \in \set{1, \ldots, n}.\> & \\
      & \quad \quad P_i \subseteq Q_i \land {} & \tag*{T3c.} \\
      & \quad \quad G_0(Q_i) \subseteq Q_i \land {} & \tag*{T3d.} \\
      & \quad \quad Q_i \times \denote{c}(P_i)(Q_i) \subseteq G_0 & \tag*{T3e.}
    \end{flalign*}
    \reqnomode
    where $I$ is viewed as a set of database states.
\end{itemize}

\begin{claim}\label{clm:0-isafety-implies-istrengthstar}
  0-\isafety $\implies$ \istrengthstar.
\end{claim}
\begin{proof}
  Consider an arbitrary set of \imp{} transactions $T$ that is 0-\isafe{} with
  respect to invariant $I$. We show $T$ is \istrongstar{} with respect to $I$.
  Let $G_0 = I \times I$. S2 holds trivially: $G_0(I) = I
  \subseteq I$. Consider an arbitrary $c$, $D$, and $D'$ where $D \in I$ and
  $(D, D') \in G_0^*$.
  \begin{align*}
    & D \in I \land (D, D') \in G_0^*  \\
    &\implies D \in I \land D' \in I \tag{Definition of $G_0$} \\
    &\implies D' \in I \land D' \circ \denote{c}(D) \in I \tag{0-\isafety} \\
    &\implies (D', D' \circ \denote{c}(D)) \in G_0 \tag{Definition of $G_0$}
  \end{align*}
\end{proof}

\begin{claim}\label{clm:istrengthstar-not-implies-0-isafety}
  \istrengthstar $\centernot\implies$ 0-\isafety
\end{claim}
\begin{proof}
  Consider a database over a single variable $x$. For convenience, we'll denote
  database states as integers. Let
  \begin{itemize}
    \item
      $G \defeq \set{(0, 0), (0, 2)} \cup \setst{(a, b)}{a \geq 2, b \geq a}$
    \item
      $I \defeq x = 0 \lor x \geq 2$, and
    \item
      the single \imp{} transaction $c \defeq
      \impif{read(x)=0}{add(x, 0)}{add(x, 1)}$.
  \end{itemize}

@@

\begin{claim}\label{clm:istrengthstar-implies-ipreservation}
  \istrengthstar $\implies$ \ipreservation
\end{claim}
\begin{proof}
  Consider a set of transactions $T$ that is \istrongstar{} with respect to
  invariant $I$ with relation $G_0$. Consider an arbitrary database $D$ that
  satisfies $I$ and an arbitrary \imp{} transaction $c$. $(D, D) \in G_0^*$,
  so by S3, $(D, D \circ \denote{c}(D)) \in G_0$. By S2, $I(D \circ \denote{c}(D))$.
\end{proof}

\begin{claim}\label{clm:istrengthstar-implies-iconfluence}
  \istrengthstar{} $\implies$ \iconfluence{}.
\end{claim}
\begin{proof}
  Consider a set of transactions $T$ that is \istrongstar{} with respect to an
  invariant $I$ with relation $G_0$. Consider an arbitrary database state $D$,
  two \imp{} chains $B^c = c_1, c_2, \ldots, c_m$ and $B^d = d_1, d_2, \ldots,
  d_n$, and two fixed transaction chains $C^c$ and $C^D$ such that $D$
  satisfies $I$, $B^c$ satisfies $I$ starting at $D$, $B^d$ satisfies $I$
  starting at $D$, $C^c = \denote{B^c}(D)$, and $C^d = \denote{B^d}(D)$. We
  show $D \circ C^c \circ C^d$ satisfies $I$.

  Let $C^z_j$ denote the prefix of $C^z$ of length $j$. We show by strong
  mathematical induction on $0 \leq i \leq m + n$, that for all natural numbers
  $j, k, j', k'$ if
    $j \leq m$,
    $k \leq n$,
    $i = j + k$,
    $j' \leq j$, and
    $k' \leq k$, then
  $(D \circ C^c_{j'} \circ C^d_{k'}, D \circ C^c_j \circ C^d_k) \in G_0^*$.

  This proof is similar to the proof of \thmref{one-is-enough}, and we can
  visualize it similar to how we visualized \thmref{one-is-enough} in
  \figref{one-is-enough}. Intuitively, our induction progresses diagonally from
  the upper left corner $D$ to the bottom right corner $D \circ C^c_j \circ
  C^d_k$. For each diagonal, our inductive hypothesis says all states on the
  diagonal are reachable via $G_0^*$ from points above and to the left.

  %
  The base case, when $i = 0$, is trivial. $i = j = k = j' = k' = 0$, so $(D
  \circ C^c_0 \circ C^d_0, D \circ C^c_0 \circ C^d_0) = (D, D) \in G_0^*$ by
  the definition of a reflexive transaction closure.
  %
  For the inductive step, consider an arbitrary set of natural numbers
  $i, j, k, j', k'$ where
    $0 < i \leq m + n$,
    $j \leq m$,
    $k \leq n$,
    $i = j + k$,
    $j' \leq j$
    $k' \leq k$.
  If $j' = j$ and $k' = k$, then $(D \circ C^c_{j'} \circ C^d_{k'}, D \circ
  C^c_{j} \circ C^d_{j}) \in G_0^*$ by the definition of a reflexive transitive
  closure. Otherwise $j' < j$ or $k' < k$. Without loss of generality, assume
  $j' < j$; the proof is symmetric for $k' < k$. We can deduce the following
  facts:
  \begin{gather}
    (D, D \circ C^c_{j-1} \circ C^d_{k}) \in G_0^*
      \label{strength-confluence-a} \\
    (D \circ C^c_{j'} \circ C^d_{k'}, D \circ C^c_{j-1} \circ C^d_{k}) \in G_0^*
      \label{strength-confluence-b} \\
    I(D \circ C^c_{j-1} \circ C^d_{k})
      \label{strength-confluence-c} \\
    (D \circ C^c_{j-1} \circ C^d_{k}, D \circ C^c_{j-1} \circ C^d_{k}) \in G_0^*
      \label{strength-confluence-d} \\
    (D \circ C^c_{j-1} \circ C^d_{k}, D \circ C^c_{j} \circ C^d_{k}) \in G_0
      \label{strength-confluence-e}
  \end{gather}
  where
  \begin{itemize}
    \item
      \eqref{strength-confluence-a} and \eqref{strength-confluence-b} are
      deduced from the inductive hypothesis applied to $i - 1$;
    \item
      \eqref{strength-confluence-c} is deduced from
      \eqref{strength-confluence-a}, the assumption $I(D)$, and S2;
    \item
      \eqref{strength-confluence-d} is deduced from the definition of a
      reflexive transitive closure;
    \item
      \eqref{strength-confluence-e} is deduced from
      \eqref{strength-confluence-d} and S3 applied to $C^c_j$, $D \circ
      C^c_{j-1} \circ C^d_{k}$, and $D \circ C^c_{j-1} \circ C^d_{k}$.
  \end{itemize}
  \eqref{strength-confluence-b} and \eqref{strength-confluence-e} imply
  $(D \circ C^c_{j'} \circ C^d_{k'}, D \circ C^c_j \circ C^d_k) \in G_0^*$.

  Finally, $I(D)$, $(D, D \circ D^c_m \circ D^d_n) \in G_0^*$, and S2 imply
  that $I(D \circ D^c_m \circ D^d_n)$.
\end{proof}

\begin{claim}\label{clm:iconfluence-not-implies-istrengthstar}
  \iconfluence{} $\centernot\implies$ \istrengthstar{}.
\end{claim}
\begin{proof}
  Consider a database over a single variable $x$. For convenience, we'll denote
  database states as integers. Let $I \defeq x = 0$ and the single \imp{}
  transaction $c = add(x, 1)$. Note that $c$ is vacuously \iconfluent{} with
  respect to $I$. Also note that $c$ is not \ipreserving{} with respect to $I$:
  $0$ satisfies $I$ but $0 \circ \denote{c}(0) = 1$ does not. By the
  contrapositive of \clmref{istrengthstar-implies-ipreservation}, $c$ is not
  \istrongstar{} with respect to $I$.
\end{proof}
